---
title: "Individualized Treatment Rules for Text Reminders for Upcoming Court Appearances"
author: Madison Coots and Jack Deschler
output: pdf_document
date: November 16, 2022
header-include: 
  - \usepackage{setspace}\doublespacing
  - \newcommand\E{\mathbb{E}}
  - \newcommand\V{\mathbb{V}}
  - \newcommand\cN{\mathcal{N}}
editor_options: 
  chunk_output_type: console
urlcolor: blue
---

\begin{center}
\textbf{Abstract}
\end{center}
Using data from a 2020 study exploring the effects of text messaging on increasing appearance rates for mandatory court appearances, we explore the effectiveness of various individualized treatment rules (ITRs). We develop ITRs for the binary decision of whether or not to treat (text) defendants in the study data, as well as ITRs for the multi-arm treatment decision of what kind of messaging strategy to use for each defendant. To this end, we explore the use of a number of different existing strategies for developing ITRs, including the use of decision trees, and evaluate their performance in this setting.

# Background
## Prior Work
In 2020, [Fishbane et al.](https://www.science.org/doi/10.1126/science.abb6591) published the results of an experiment conducted in New York to explore the causal effects of nudges in the criminal justice context. The authors were interested in understanding whether text message reminders regarding an individual’s upcoming mandatory court appearance could help that individual avoid a failure to appear (FTA). Ultimately, they found that providing text message reminders increased the likelihood that defendants would show up to their appointed court date, thus eliminating a substantial percentage of arrest warrants for failing to appear in court.

In addition to conducting experiments probing the effects of text reminders generally, the authors also experimented with employing different messaging strategies. Among the clients randomly selected to receive text messages, clients were again randomized across three different text message arms. Clients in the first arm received text reminders that emphasized the adverse consequences of missing a court appearance, clients in the second arm received messages that emphasized and encouraged clients to plan for their upcoming appearance, and clients in the third arm received messages that were a combination of the two previous messaging approaches.

## Motivation
In the U.S. criminal justice system, failing to appear for mandatory court appearances can result in severe consequences for the defendant. Many courts will issue a bench warrant for a defendant's arrest if they fail to appear, which means that the defendant may be taken to and held in jail the next time they come into contact with courts or law enforcement. Helping clients remember their upcoming mandatory court appointments therefore emerges as a low cost and effective method for helping people avoid pretrial incarceration and its collateral consequences on families and communities.

Given the high stakes associated with missing a court appearance, it is imperative that efforts to remind defendants of their appointments via text do not unintentionally discourage clients from appearing in court. In particular, while highlighting the negative consequences of missing court in the text reminders may motivate certain types of defendants to appear, that same framing may discourage other defendants. In particular, defendants from communities that are disproportionately impacted by policing might have a pre-existing distrust of the system, and highlighting the negative consequences of missing a court appearance may further deepen a sense of fear and distrust about engaging with the system. To address this potential heterogeneity in the way defendants may respond to different messaging strategies, an exploration of individualized treatment rules is warranted in this setting.

## Methods

For the first phase of this project, we focus on developing ITRs using causal forests, which are a learning method for causal inference that are an extension of Random Forests. Causal forests are built similarly to random forests. However, rather than repeatedly splitting the data in order to minimize prediction error of an outcome variable, causal forests split the data in order to maximize the difference between an outcome variable and a treatment variable. Ultimately, this enables us to estimate how treatment effects may vary across a sample, and develop individualized treatment rules.

To evaluate the effectiveness of an ITR, we follow [Imai and Li (2019)](https://www.tandfonline.com/doi/full/10.1080/01621459.2021.1923511) and use the population average prescriptive effect, or PAPE, as our primary performance metric. The PAPE's key advantage over metrics like the population average value is that it adjusts for a different proportion of the population being treated under an ITR than was treated in the original randomized experiment. This feature is especially useful in the current setting, where the randomized experiment treated a very small portion of the individuals who received a summons (about 6%, when restricted to only those using the new summons forms). Imai and Li (2019) propose another metric to evaluate ITRs, the Area Under the Prescriptive Effect Curve, or AUPEC. The AUPEC for an ITR is the integral of the the PAPE for all budget constraints. As such, our paper focuses on the PAPE and does not deal with AUPEC to evaluate our ITRs. Unlike the treatments considered by Imai and Li, text messages are extremely low cost, and so a budget constraint makes less sense in this particular setting. For this reason, we do not compute AUPEC.

# Data Summary

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# libraries and setup
library(evalITR)
library(dplyr)
library(haven)
library(kableExtra)

madison_path <- "/Users/madisoncoots/Documents/harvard/research/fishbane/Replication package/Field experiment data and code/TextMessageData.dta"

jack_path <- "../Replication package/Field experiment data and code/TextMessageData.dta"

# read data HAVE TO SWITCH PATHS
df <- read_dta(madison_path)

# subset to only those with the newform?
df <- df[df$NewForm == 1,]

# Create Treatment column (original does not work for some reason)
df$Treatment <- df$consequences + df$planmaking + df$combination

# Flip FTA because a success is actually appearing
df$success <- 1 - df$FTA


control <- df[df$Treatment == 0,]
treated_overall <- df[df$Treatment == 1,]
cons <- df[df$consequences == 1,]
plan <- df[df$planmaking == 1,]
combo <- df[df$combination == 1,]
```


```{r include = FALSE}
# Function to easily extract appearance rates
appearance_rate_by_demo <- function(data, treatment = "All", demo = "All", threshold = 0.5, places = 3) {
  if (demo == "All" & treatment == 'All')
    tmp <- data
  else if (demo == "All")
    tmp <- data[which(data[treatment] == 1),]
  else if  (treatment == "All")
    tmp <- data[which(data[demo] > threshold),]
  else
    tmp <- data[which(data[demo] > threshold & data[treatment] == 1),]
  return(round(1 - (sum(tmp$FTA) / nrow(tmp)), places))
}
```

The authors of Fishbane et al. released a rich dataset of their study results that inlcuded a number of informative covariates on each study participant. The covariates included demographic information for each defendant, as well as limited criminal history information (relating to past summons and past failures to appear in court), and details about the offense that resulted in the defendant's court summons.

As a preliminary step, we examine the overall appearance rates that resulted from employing one of the four text message treatment arms. These results are shown in Table 1. We observe that the combination messaging strategy yielded the highest appearance rate.

```{r echo = FALSE}

no_text <- c(nrow(control) - sum(control$FTA), sum(control$FTA), round(1 - (sum(control$FTA)/nrow(control)), 3))
planning <- c(nrow(plan) - sum(plan$FTA), sum(plan$FTA), round(1 - (sum(plan$FTA)/nrow(plan)), 3))
consequences <- c(nrow(cons) - sum(cons$FTA), sum(cons$FTA), round(1 - (sum(cons$FTA)/nrow(cons)), 3))
combination <- c(nrow(combo) - sum(combo$FTA), sum(combo$FTA), round(1 - (sum(combo$FTA)/nrow(combo)), 3))
overall <- c(nrow(df) - sum(df$FTA), sum(df$FTA), round(1 - (sum(df$FTA)/nrow(df)), 3))

data.frame(no_text,
           planning,
           consequences, 
           combination, 
           overall) %>%
  `rownames<-`(c("Appeared", "Failed to appear", "Appearance rate")) %>%
  knitr::kable(col.names = c("No text", "Planning", "Consequences", "Combination", "Overall"),
               caption = "Conditional appearance rates for each treatment arm") %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(3,bold=T) %>%
  row_spec(0,bold=T)
```

We also examine appearance rates along a number of different demographic dimensions, including gender, race, and poverty level. We note that the original dataset does not contain a variable for the race of an individual, but instead contains the percentage of the population in each individual's zip code that is either Black or Hispanic. Ideally, we would impute race by sampling a race label for each defendant according to the distribution of their zip code. However, the breakdown supplied in the data is not a distribution, so we cannot employ this method of imputing race. As a (temporary) workaround, we can construct an approximation of an individual's race by coding someone as, for example, Black if more than 50% of the population in that ZIP code was Black. We can do the same based on the percentage below the poverty line column. Table 2 below summarizes the appearance rates for various demographics in the data, using the heuristic just described.

```{r include=F}
# Demographic Breakdowns
df2 <- df
df2['Male'] = 1 - df2$Female
df2['PctBlackHispZip'] = df2$PctBlackZip + df2$PctHispZip
demos = c('Male', "Female", "PctBlackZip", "PctHispZip", "PctBlackHispZip", "PctBelowPovZip")
res <- matrix(data = 0, nrow = length(demos), ncol = 4)
for (d in  1:length(demos)) {
  res[d, 1] <- appearance_rate_by_demo(df2, demo = demos[d])
  res[d, 2] <- appearance_rate_by_demo(df2, treatment = "planmaking", demo = demos[d])
  res[d, 3] <- appearance_rate_by_demo(df2, treatment = "consequences", demo = demos[d])
  res[d, 4] <- appearance_rate_by_demo(df2, treatment = "combination", demo = demos[d])
}
```

```{r echo = FALSE}
demographic <- c("Gender", "Gender", "Race", "Race", "Race", "Poverty")
group <- c("Male", "Female", "Black", "Hispanic", "Black or Hispanic", "Below Poverty Line")
overall <- c(res[1,1], res[2,1], res[3,1], res[4,1], res[5,1], res[6,1])
planmaking <- c(res[1,2], res[2,2], res[3,2], res[4,2], res[5,2], res[6,2])
consequences <- c(res[1,3], res[2,3], res[3,3], res[4,3], res[5,3], res[6,3])
combination <- c(res[1,4], res[2,4], res[3,4], res[4,4], res[5,4], res[6,4])

data.frame(demographic,
           group,
           overall,
           planmaking,
           consequences,
           combination) %>%
  knitr::kable(col.names = c("Demographic",
                             "Group",
                             "Overall",
                             "Planmaking",
                             "Consequences",
                             "Combination"),
               caption = "Conditional appearance rates for each treatment arm for different demographic groups") %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0,bold=T)
```

\smallskip

We likewise break down the appearance rate for each of the treatment arms by NYC borough. These results are shown below in Table 3. Interestingly, we observe that residents of Staten Island experience a higher appearance rate in response to the consequence-focused text reminders than do residents of other boroughs in New York.

```{r include=F}
# Demographic Breakdowns
df2 <- df
demos = c('Bronx', "Brooklyn", "Manhattan", "Queens", "StatenIsland")
res <- matrix(data = 0, nrow = length(demos), ncol = 5)
for (d in  1:length(demos)) {
  res[d, 1] <- appearance_rate_by_demo(df2, demo = demos[d])
  res[d, 2] <- appearance_rate_by_demo(df2, treatment = "Treatment", demo = demos[d])
  res[d, 3] <- appearance_rate_by_demo(df2, treatment = "planmaking", demo = demos[d])
  res[d, 4] <- appearance_rate_by_demo(df2, treatment = "consequences", demo = demos[d])
  res[d, 5] <- appearance_rate_by_demo(df2, treatment = "combination", demo = demos[d])
}
```

```{r echo = FALSE}
boroughs <- c("The Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island")
overall <- c(res[1,2], res[2,2], res[3,2], res[4,2], res[5,2])
any_treatment <- c(res[1,1], res[2,1], res[3,1], res[4,1], res[5,1])
planmaking <- c(res[1,3], res[2,3], res[3,3], res[4,3], res[5,3])
consequences <- c(res[1,4], res[2,4], res[3,4], res[4,4], res[5,4])
combination <- c(res[1,5], res[2,5], res[3,5], res[4,5], res[5,5])

data.frame(boroughs,
           any_treatment,
           overall,
           planmaking,
           consequences,
           combination) %>%
  knitr::kable(col.names = c("Borough",
                             "Overall",
                             "Any treatment",
                             "Planmaking",
                             "Consequences",
                             "Combination"),
               caption = "Conditional appearance rates for each treatment arm by borough") %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0,bold=T)
```

# Preliminary Results

## Binary Causal Forest

We fit causal forests of three different sizes (100, 500, and 1000 trees) to generate individualized treatment rules for the binary treatment decision (text reminders of any kind, or no texts). For each forest, we also experimented with three different sets of covariates: offense-based, defendant-based, and a combination of these two sets. The first set contains covariates characterizing the offense that resulted in the court summons, the second set characterizes the defendant in terms of demographics and criminal history, and the third is simply the union of these first two sets. We list the comprising covariates of each set below:

- Offense-based: 1) borough of the offense, 2) whether the offense took place in the park after dark, 3) whether the offense involved alcohol or 4) marijuana, 5) whether the offense involved passive or 6) active disorderly conduct, 7) whether the offense was biking on the sidewalk, 8) unreasonable noise, 9) reckless driving, or 10) public urination.
- Defendant-based: 1) gender, 2) age, 3) whether the person has failed to appear in court in the past, 4) the number of past failures to appear, 5) whether the person has received a court summons in the past, 6) the number of past court summons, 7) the median income of their residential zip code, 8) the percentage of Black and 9) Hispanic residents in the defendant's residential zip code, and 10) the population percentage that is below the poverty line in the defendant's zip code

After fitting the causal forests to the study data, we also used two different methods of generating individualized treatment rules from the resulting models: simple, and strict. To implement the simple treatment rule, we simply examine the treatment effect given by the causal forets for each defendant in the data and treat them if it is greater than 0. Formalized, this treatment rule is $\hat\tau_i > 0$. To implement the strict treatment rule, we reason that to ensure that we do no harm to the defendant by texting them, a 95\% confidence interval around the individual treatment effect should \textit{not} contain 0. If this is true, then the unit receives treatment. This stricter, "do no harm" treatment rule can be formalized as $Pr(\mathcal{N}(\hat\tau_i, \hat\sigma_i^2) \leq 0) < .025$.

In the table below, we show the PAPE its standard deviation of each of the ITRs generated by the methodology described above. From these results, we see that the causal forests trained using the offense-based covariates with the strict method are the highest performing.

```{r echo = FALSE}
results_path <- "binaryITR_pape.csv"
binary_itr_results <- read.csv(results_path)

binary_itr_results %>%
  mutate(covariates = case_when(covariates == "Offense" ~ "Offense-based",
                                covariates == "Perpetrator" ~ "Defendant-based",
                                TRUE ~ covariates),
         rule = case_when(rule == "simple" ~ "Simple",
                          rule == "harmless" ~ "Strict",
                          TRUE ~ rule),
         papes = round(papes, digits = 4),
         sds = round(sds, digits = 4)) %>%
  select(numtrees, covariates, rule, papes, sds) %>%
  arrange(desc(papes)) %>%
  knitr::kable(col.names = c("No. of trees", "Covariate Set", "Treatment Rule", "PAPE", "Standard dev."),
               caption = "Binary Causal Forest ITR PAPEs") %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0,bold=T)
```

Interestingly, these results indicate that the PAPE is not dependent upon the number of trees with which a causal forest is fit. We will check whether this is due to a coding error, but if it is true, these results indicate that we can likely fit forests on smaller numbers of trees, thus reducing computation time. The results also indicate that an offense-based set of covariates with a strict rule is the best way to improve over the random experiment with an ITR.

## Multi-arm Causal Forest

We also experimented with fitting a causal forest to generate a treatment rule for the multiarm version of the experiment: no texts, consequence-focused texts, planning-focused texts, or combination texts. Due to computational limitations we show the results for one causal forest only: 100 trees using all covariates and the strict method for generating the ITRs. Furthermore, because this causal forest generates a multiarm treatment rule, the output is best interpreted through a matrix, where the interpretation of a given entry is: the PAPE when comparing one specific ITR treatment to a specific randomized experiment treatment condition. Below, we show such a matrix of PAPEs for the multi-arm causal forest and its corresponding matrix of standard deviations of the PAPEs. The columns correspond to ITR treatments, and the rows to experiment treatments. 

From this result, we can see that, when compared to a purely randomized experiment, the ITR for the consequences text messaging strategy achieves the highest PAPE value ($\approx 0.0135$).

```{r echo = FALSE}

row_names <- c("Any text", "Planning", "Consequences", "Combo")

pape_matrix <- data.frame(any_text = c(0.01047226, 0.01150315, 0.0112156, 0.004948374),
                          planning = c(-0.004180083, 0.001993335, -0.003877044, -0.01600147),
                          consquences = c(0.005797335, -0.0001260121, 0.01348087, 0.000954789),
                          combo = c(0.01090064, 0.01320184, 0.007252543, 0.0115389)) %>%
  `rownames<-`(row_names)

sd_matrix <- data.frame(any_text = c(0.003793225, 0.005854321, 0.005913416, 0.008414251),
                          planning = c(0.002900054, 0.004554933, 0.004522923, 0.006186608),
                          consquences = c(0.003466761, 0.005312291, 0.005462719, 0.007636758),
                          combo = c(0.003105024, 0.004833518, 0.004808774, 0.006907275)) %>%
  `rownames<-`(row_names)

pape_matrix %>%
  knitr::kable(col.names = row_names,
               caption = "Multiarm Causal Forest ITR PAPEs") %>%
  kable_styling(latex_options = "HOLD_position")

sd_matrix %>%
  knitr::kable(col.names = row_names,
               caption = "Multiarm Causal Forest ITR Standard Deviations") %>%
  kable_styling(latex_options = "HOLD_position")

```

# Future Work
Looking forward to the second phase of this project, there are several extensions we are interested in further exploring. On the evaluation side, we hope to think about how to compare the ITRs that result from the various causal forests we examine. Specifically, we are interested in thinking about whether comparing PAPEs across forests is most appropriate, or if there are other metrics and factors worth considering. Along similar lines, we also hope to think about ways to generate a single performance metric for multiarm causal forests from the matrix of PAPEs that we currently generate for each forest. 

We are also interested in exploring the use of other algorithms for generating ITRs. In addition to acquiring results for multiarm causal forests of additional sizes, we plan on exploring the use of support vector machines (SVMs) for ITRs. Finally, we are also intrigued by differences in appearance rates that we observe across different boroughs in response to the various treatment arms. To this end, we plan on conducting a permutation test to see if it is possible to attain a high-performance ITR using a simple rules-based approach. This exploration, in particular, could yield valuable policy insights regarding the quality of ITRs generated with interpretability and simplicity as a priority. 
